Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-08 paper.tex
update figures

@@

\section{Introduction}
% What are ARGs, their relevance and importance

@@

% previous likelihood calculations on ARGsL Kuhner and Felsenstein (2000), ARGweaver

% history of approximating the coalescent: 
SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
(2000)
These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
Clear distinction into two groups: ARGWeaver infers actual ARG.
% what does kwarg do?
Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
structures. (other peculiarities: ...).

@@

\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
Recording unary nodes has interesting consequences on the left-to-right 
description of the SMC. Looking at a single marginal tree, the presence of 
unary nodes tells us that either \\

@@

\begin{figure}[h!] \label{fig:algo-flat}
\centering
\includegraphics[width=\textwidth]{figures/smc_algo_flat.png}
\caption{Description of algorithm}
\end{figure}

@@

% improve this section, transition to modifications in the algorithm in case of 
% incomplete unary nodes, polytomies, ...
We further rely on the assumption that the unary (coalescent) nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm

% keep track of last parent to identify recombination events
% recombination events beyond root: we keep track of
% parent of prior tree

% extent of lineages

% version of the algorithm with polytomies

% version of the algorithm in the absence of unary nodes

@@

% highlight in what ways we can now compute likelihoods for a GENERAL class of ARGs

@@

Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample.

% something on the process backwards in time
We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).

% HOW DO WE HIGHLIGHT THESE FEATURES: SUPPLEMENTS?
\subsubsection{Slicing the ARG in time}
% TO DO: give example why this might be useful.

\subsubsection{Slicing the ARG in (genome) space}
%TO DO: give example why this might be useful.

\subsubsection{Variable Ne}

\subsubsection{Variable r}

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. 
Important step moving forwards towards bridging the gap between the tremendous 
recent advances in ARG inference and current state of the art statistical inference of 
past demography and selection (but see ). % some/many of these approaches

% ignore recombination

% future directions: 
incorporating node labels (see \citep{guo_recombination-aware_2022}).

First step towards a genealogy sampler (for any ARG, and at scale): 
quantifying uncertainty 
(with inferred ARG as starting point). Compared to existing MCMC approaches, 
we further restrict the size of ARG space by 
integrating out the exact time to the recombination event.

% discuss accurracy:
% there will be scenarios where the assumptions made here are not met: e.g. polytomies
% yet we can still always compute a likelihood.
% we can alwyas associate a value, what remains to be verified what the information is in that
% case.

% explore the information in unary nodes

@@

\section{Supplementary material}

>>>>

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% deal with supplementary items
\newcommand{\supplementarysection}{%
  \setcounter{figure}{0}% Reset figure counter
  \let\oldthefigure\thefigure% Capture figure numbering scheme
  \renewcommand{\thefigure}{S\oldthefigure}% Prefix figure number with S
  \section{Supplementary section}% Set supplementary section
}

@@

\section{Introduction}

% Main motivation:
Major progress in ARG inference.

Result is whole range of inference methods, all inferring an ARG with their own specific 
characteristics.
Many of these scalable heuristic methods do not explicitly infer recombination events.

%ARGs that are inferred here lack information to enable us to compute 
% a likelihood under the Hudson coalescent.
% Missing: exact time to recombination + non-inferrable recombinations

Currently no way to connect the output of the diverse range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, (or likelihood-based inference, ...)

% What are ARGs, their relevance and importance

@@

% previous likelihood calculations on ARGs Kuhner and Felsenstein (2000), ARGweaver

% history of approximating the coalescent: 
%SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
%Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly (2000)
%These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
%to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
%tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
%Clear distinction into two groups: ARGWeaver infers actual ARG.
% what does kwarg do?
%Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
%the built in assumption of neighbouring trees being separated by single Subtree Prune 
%Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
%structures. (other peculiarities: ...).

@@

\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
Recording unary nodes has interesting consequences on the left-to-right 
description of the SMC. Looking at a single marginal tree, the presence of 
unary nodes are telltale signs of future or past coalescence events (in a 
left-to-right sense).\\

@@

\begin{figure}[h!] \label{fig:algo-flat}
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from the first 
local tree (1) to the next (4). For each edge that is removed from the first tree we 
decrement the count in the corresponding internode intervals and update the identity of 
the last parent. Subsequently, for each edge that will be inserted we compute the likelihood 
associated with that edge prior to incrementing the counts.}
\end{figure}

@@

% ADD IN RESULTS
Figure \ref{sup:fig:vs-hudson} shows the correlation between the likelihood under the hudson 
for an ARG with detailed recombination information simulated and computed using 
\texttt{msprime} \citep{baumdicker_efficient_2021}. Show impact of unary nodes on $r^2$.
Simulated mutation data under the SMC and passed this to ARGweaver.
Scatterplot of coalescent prior as returned by ARGweaver \citep{rasmussen_genome-wide_2014} 
for a 100 samples from the posterior against the likelihood as described here fore these 
sampled ARGs \ref{sup:fig:vs-argweaver}

Implementation of the algorithm is available on https://github.com/GertjanBisschop/runsmc.

@@

@@

Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome.

% Algorithm is suitable for a general class of ARGs each with their own specific features.
% What happens in the case of ARGs that do not meet our assumptions: absent unary nodes, polytomies, ...
The above algorithm treats each edge independently which implies we can deal with polytomies.
Furthermore, although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm can also accomodate for 
those inferred ARGs that lack unary nodes (e.g.\ \relate, and see figure in supplements).
However, in case of (sub)tree height changes along the ARG, the algorithm would infer 
two recombination events along the lineages associated 
with both edges switching parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1 (see supplementary figure).

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs.

Impact both on ways of doing ARG inference itself and quantifying the uncertainty associated 
with that as well as performing statistical inference of past demography.

* statistical inference *
This is an important step towards bridging the gap between the tremendous 
recent advances in ARG inference and the current state of the art statistical inference of 
past demography (but see ). 
% some/many of these approaches only use the information present in marginal trees.
The provided theory and associated algorithm will function as a building block of a 
demography-aware genealogy sampler. Current approach couple of major advantages:
* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
MCMC runtime by reducing the necessary burnin period.
* by integrating out the exact time to the recombination event we further restrict the size 
of ARG space that needs to be explored.
* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
* we can readily incorporate $N_e$-changes as well as multiple populations and migration.

* ARG inference *
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}
% tsinfer not model-based, minimal assumptions, heuristic, scalable

quantify uncertainty
* polytomies represent uncertainty:
*  systematicallty break them by sampling from the coalescent rather then randomly
* associate uncertainty as metadata to parent node representing uncertainty in the
* interval spanned by the polytomy
* can something similar be done for stacked recombinations?

@@

\pagebreak

\supplementarysection
\section*{Supplementary Information}

\begin{figure}[h!] \label{sup:fig:vs-argweaver}
\centering
\includegraphics[width=\textwidth]{example-image-a}
\caption{Compare likelihoods inferred ARGs ARGweaver}
\end{figure}

\begin{figure}[h!] \label{sup:fig:vs-hudson}
\centering
\includegraphics[width=\textwidth]{example-image-a}
\caption{Compare likelihoods Hudson and SMC}
\end{figure}
